\chapter{Acknowledgements}

\onehalfspacing

I would like to take this opportunity to extend my most heartfelt gratitude to
my guide and mentor - \textbf{Prof. Manoj Misra}, Department of Electronics and
Computer Engineering, IIT Roorkee. His guidance, attention to detail and drive
for excellence have contributed greatly towards my education and also towards
ensuring that sterling work is done as part of my Masters dissertation. His
encouragement and unstinting faith have been essential in helping me learn, grow
and push the limits of my capabilities. I am fortunate to have been able to do
my dissertation under him.

My sincere thanks are due to \textbf{Dr. S. N. Sinha}, Professor and Head of the
Department of Electronics and Computer Engineering for providing the facilities
and hardware without which this thesis would have been all but impossible.
Thanks are also due to \textbf{Dr. Raheja}, Institute Architect, IIT Roorkee for
providing the blueprints of Ravindra Bhawan, the test site for this thesis. My
work would have been immensely more difficult without his help and cooperation.
The \textbf{Mahatma Gandhi Central Library} of IIT Roorkee also deserves a
mention in the Acknowledgements. They have provided unfettered access to the
world's highest quality research published in the IEEE and ACM journals. I have
always felt well supplied with information thanks to their hard work.

Sincere gratitude is also due to the authors of the many open source software
projects that have been very useful in the making of this thesis. The list of
projects is very long and it is impossible to name all, but I would like to give
a special mention to the authors and contributors of \textbf{\LaTeX, GNUPlot,
the GNU R Language, the GNU Core Utilities, Python, Linux, Android} and
\textbf{Eclipse}. This thesis is the result of standing on the shoulders of
giants and I am indebted to you for your efforts in maintaining such high
quality tools for the benefit of the world. Many many thanks.

Acknowledgements aren't complete without thanking \textbf{friends and family}.
They are the ones who have given unconditional encouragement and support. They
have shared my happiness on the days that things have worked and they have
helped get me through the days when things haven't. They have stood by me
through thick and thin and for that I am very grateful.

