\chapter{Introduction}

\section{Indoor Positioning}

Indoor positioning has been a topic of active research for the past decade with the first research system using RF signals for distance estimation and indoor positioning being produced by Microsoft Research (RADAR, 2000)\cite{RADAR}. Of course, systems like Active Badge  have been developed as far back as 1992 to locate targets indoors but they have shown limited utility on account of the requirement to deploy specialized sensors for detecting the radio tags deployed within the system.

Since RADAR, a number of improvements have been made in the core algorithms and technologies for positioning. Over time, indoor positioning has progressed from simply using RF signals to using ubiquitous 802.11 Access Points as radio sources. With the development of UWB (Ultra Wide Band) technology, very fine grained ranging and tracking results have been obtained over target distances of the order of a few hundred feet from the radio source. 

However, the primary scale of interest for commercial exploitation of indoor positioning and tracking is of the order of the size of warehouses and malls (~2500 m2 or more) with an accuracy that is preferably of the order of a few meters. The target applications for such a scenario are inventory location and tracking; location based services and personalized offers from businesses based on physical proximity to their shops or other points of sale. The ultimate target to be achieved in this line of research is:

Create a low complexity, low cost indoor positioning system that is maintainable, reasonably accurate and highly scalable.

UWB, unfortunately, does not scale easily to such large spaces without requiring that a significant number of sensors and repeaters be installed in the target positioning area specifically for this purpose. It is also an expensive technology. On the other hand, IEEE 802.11 AP based positioning has low complexity and cost and it has a large installed base, thus making it attractive for positioning purposes as a source of radio signals.

Thus it is interesting to explore the limitations of accuracy, scalability and maintainability of indoor positioning systems that are based on IEEE 802.11 Access Points.


\section{Indoor Tracking}

While exploring indoor positioning, another major application of such a system leaps out - Indoor Tracking. Indoor tracking refers to the process of tracking the location of a body on a continuous basis.


\section{Motivation}

  \subsection{Indoor Navigation}

  Describe indoor navigation applications (shops etc.)
  Why is it required?

  \subsubsection{Shipping industry}
  Courier delivery etc.


  \subsection{Augmented Reality}

  Describe AR.

  \subsubsection{Fusion Gaming}

  Describe potential applications to games in the real world.

  Describe AR, real-world games, indoor navigation, shopping complexes, warehouses.


\section{Problem Statement}
  Describe the problem to be solved.